\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-07-07 ResponseToReview.tex

3) Why are sample sizes not reported?  You provide total samples sizes (eg, 206 or 216 for Phenotyping 1, depending on whether you read the Methods or the Results), but not sample sizes for each genotypic class.\\

\res{This has been addressed. We report accurate sample sizes for both phenotypings that are consistent throughout the text.}

4) It was not clear to me what “to detect the observed effect” means in lines 12-13 of p. 9.  Your Phenotyping 1 results show no effect of genotype on tillering.  Thus, it is not clear what effect you are referring to.\\

\res{We have clarified this in the text: "Based on these initial data, we conducted a \emph{post hoc} power analysis using effect size data for \emph{tb1} associated QTL from (Briggs \emph{et al.} 2007), which indicated that a minimum of 71 individuals in each genotypic class would be needed to detect the suggested effect of the \emph{Hopscotch} on tillering index from Briggs \emph{et al.} 2007."}

5) What was the purpose of the Phenotyping 2  experiment?  You already showed no effect of genotype on phenotype using populations where you are most likely to see segregation for the insertion, so it was not clear to me why sampling more broadly, especially from populations with no evidence of the Hopscotch insertion, would be expected to improve your understanding of this relationship.\\

\res{You are correct that our Phenotyping 1 did not show an effect of the Hopscotch on tillering index; however, the distribution of genotype classes was so heavily skewed to individuals with the Hopscotch insertion (both heterozygotes and homozygotes) that we felt we may not have had enough individuals without the insertion to have been able to detect an effect.}

\res{Our sampling for phenotyping experiments was limited to a population from which we had a large number of seed collected, and we were unable to survey more than one population for these experiments. For our more broad genotyping of both landrace maize, ssp. \emph{parviglumis}, and ssp. \emph{mexicana} we sampled as many population as we could. For most of these populations the prevalence of the Hopscotch was previously unknown, and we were interested in having an idea of the overall distribution and frequency of the Hopscotch throughout these locations.}

6) The discussion of introgression, genetic drift, and selection in your Discussion section seems to lack coherence.  It sometimes focuses on explaining the unexpectedly high frequency of Hopscotch in present-day populations, whereas other times it seems to be addressing the presence or absence of Hopscotch in teosinte in general, and it also sometimes seems to be addressing selection on the tb1 locus.

### 2014-07-07 ResponseToReview.tex

\section*{Reviewer 1}

%Reviewer \#1: In their paper 'Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)', Vann et al survey a large sample of maize and teosinte individuals for the presence of the Hopscotch transposable element upstream of tb1. The authors find that Hopscotch is more widespread, across a large sampled area in Mexico, than previously thought, particularly in parviglumis. The authors go on to sequence regions up and downstream of Hopscotch in subset of their initial sample. In analyzing their sequence data, they find no evidence for recent introgression of the maize Hopscotch locus into wild teosintes. They also find evidence for selection acting on tb1 in parviglumis. In greenhouse experiments, no difference in tiller index or tiller number was observed between teosinte lines with or without Hopscotch.

%\res{yes sounds right}

1) They present an interesting result, their conclusions are supported by their data, and the methods they use are appropriate. However, I feel like they need to make some further effort to understand what is happening mechanistically. What is Hopscotch doing in maize vs these sampled teosintes? I feel like the expression analyses and phenotyping experiments they suggest should be part of a later study should be part of this study. At a minimum, qRT-PCR experiments should be performed assessing tb1 expression in their with/without Hopscotch lines. Expression analyses of gt1, te1, and tru1 would be great, since they invoke variation at these loci to explain their results.

Another potential avenue, that the authors also suggest, lies in more directed and exhaustive greenhouse experiments. Variation in red and far-red light has been shown to be important in regulating both tb1 (in Sorghum) and gt1 in maize. Growth chamber experiments (with or without supplemental far red light, with or without Hopscotch), coupled to expression analyses in seedlings, seem like they should be possible.  In the greenhouse experiments that were done, phenotyping was incomplete. Apart from tillering, Hopscotch and natural variation at the tb1 locus appears to affect ear architecture and internode length. Was there any evidence for variation in these other traits? In the longer term, it would be nice to see what happens when you introgress these particular parviglumis loci into maize.\\

\res{Expression analyses of \emph{tb1} and additional loci coupled with a more exhaustive phenotyping would be interesting and useful in understanding the role of \emph{tb1} in branching architecture in teosinte, but is beyond the scope of our study. The suggested experiments would in themselves be enough for an additional paper. While we are unable to include these experiments in this paper we have added text to the conclusions of this paper discussing their utility and the need for further work to elucidate the role of \emph{tb1} in branching architecture in teosinte.}

2) One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.

### 2014-07-07 ResponseToReview.tex

%\res{yes sounds right}

1) They present an interesting result, their conclusions are supported by their data, and the methods they use are appropriate. However, I feel like they need to make some further effort to understand what is happening mechanistically. What is Hopscotch doing in maize vs these sampled teosintes? I feel like the expression analyses and phenotyping experiments they suggest should be part of a later study should be part of this study. At a minimum, qRT-PCR experiments should be performed assessing tb1 expression in their with/without Hopscotch lines. Expression analyses of gt1, te1, and tru1 would be great, since they invoke variation at these loci to explain their results.

Another potential avenue, that the authors also suggest, lies in more directed and exhaustive greenhouse experiments. Variation in red and far-red light has been shown to be important in regulating both tb1 (in Sorghum) and gt1 in maize. Growth chamber experiments (with or without supplemental far red light, with or without Hopscotch), coupled to expression analyses in seedlings, seem like they should be possible.  In the greenhouse experiments that were done, phenotyping was incomplete. Apart from tillering, Hopscotch and natural variation at the tb1 locus appears to affect ear architecture and internode length. Was there any evidence for variation in these other traits? In the longer term, it would be nice to see what happens when you introgress these particular parviglumis loci into maize.\\

\res{Expression analyses of \emph{tb1} and additional loci coupled with a more exhaustive phenotyping would be interesting and useful in understanding the role of \emph{tb1} in branching architecture in teosinte, but is beyond the scope of our study. The suggested experiments would in themselves be enough for an additional paper. While we are unable to include these experiments in this paper we have text to the conclusions of this paper discussing their utility and the need for further work to elucidate the role of \emph{tb1} in branching architecture in teosinte: "Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}) as well as include a more exhaustive phenotyping including all traits.  These data, in conjunction with more exhaustive phenotyping, should help reveal the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte."}

### 2014-07-08 ResponseToReview.tex

\res{Expression analyses of \emph{tb1} and additional loci coupled with a more exhaustive phenotyping would be interesting and useful in understanding the role of \emph{tb1} in branching architecture in teosinte, but is beyond the scope of our study. The suggested experiments would in themselves be enough for an additional paper. While we are unable to include these experiments in this paper we have text to the conclusions of this paper discussing their utility and the need for further work to elucidate the role of \emph{tb1} in branching architecture in teosinte: "Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}) as well as include a more exhaustive phenotyping including all traits.  These data, in conjunction with more exhaustive phenotyping, should help reveal the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte."}

2) One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.

\res{is this necessary?}
\jri{add to figure}

3) Reviewer 2: Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.\\

\jri{explain point. add minor text to paper clarifying distinction, provide text}

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and some landrace maize; however, previous studies did not sample as many populations or individuals per population as the study we present. Our study is not only concerned with whether or not the Hopscotch is present or absent in these populations but is also interested in what its frequency is so that we can better understand the evolutionary history of this insertion in teosinte.}

### 2014-07-09 ResponseToReview.tex

2) One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.

\res{is this necessary?}
\jri{add to figure}

3) Reviewer 2: Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.\\

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and some landrace maize; however, previous studies did not sample as many populations or individuals per population as the study we present here. Our study is not only concerned with whether or not the \emph{Hopscotch} is present or absent in these populations but is also interested in what its frequency is so that we can better understand the evolutionary history of this insertion in teosinte. We have added clarifying text to the introduction: "The effects of the \emph{Hopscotch} insertion have been studied in maize (Studer \emph{et al.} 2011), and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes (Studer and Doebley 2012), but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and landrace maize; however little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations."}

4) Nevertheless, the manuscript is worth publishing as it does present data that may help rule out some possible explanations as to why tb1 is so common in parviglumus.\\

\res{Thank you; we hope that our paper in conjunction with future research and experiments can help to further the understanding of the role of both \emph{tb1} and the \emph{Hopscotch} insertion in teosinte.}

5) This paper is fairly maize-centric. A more general message is included in the discussion, not as much in the introduction, and not at all in the abstract. This could be easily remedied.\\

\jri{tweak intro and abstract, add a bit to discussion maybe, explain text changes here.}
\res{We have added sentences to the abstract and discussion in order to make 
Methods and statistics were appropriately used as far as my expertise allows me to judge.

### 2014-07-09 manuscript.tex
edits
additional experiments discussed
Reviewer 2 edits
Addressing reviewer 2 edits

%TO DO
%check citations in last paragraph for te1

%preprint somewhere
%finish fixing paragraph that jeff doesn't like
%check map details
%Code for SAS and command line stuff on Git
%Phenotyping on Figshare http://dx.doi.org/10.6084/m9.figshare.776926 -- do we need to cite this in the manuscript somewhere?

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte.

@@

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that the insertion would have drifted to high frequency in multiple populations. We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}; this coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations suggests that the locus may play an ecological role in teosinte. In contrast to domesticated maize, the \emph{Hopscotch} insertion does not appear to have a large effect on tillering in \emph{parviglumis}. Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}).  These data, in conjunction with more exhaustive phenotyping, should help reveal the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte.

>>>>

%Code for SAS and command line stuff on Git
%Phenotyping on Figshare http://dx.doi.org/10.6084/m9.figshare.776926 -- do we need to cite this in the manuscript somewhere?

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and landrace maize; however little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations.

@@

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that the insertion would have drifted to high frequency in multiple populations. We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}; this coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations suggests that the locus may play an ecological role in teosinte. In contrast to domesticated maize, the \emph{Hopscotch} insertion does not appear to have a large effect on tillering in \emph{parviglumis}. Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}) as well as include a more exhaustive phenotyping including all traits.  These data, in conjunction with more exhaustive phenotyping, should help reveal the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte.
